Web Services Composition in the presence of Non-Functional Properties: A MDD Approach

### 2013-09-30 WESOA2013/source/paper.tex
inicial --martin

\chapter{Evaluation}
\label{chapter:validation}

\epigraph{``\textit{Let us change our traditional attitude to the construction
of programs: Instead of imagining that our main task is to instruct a computer
what to do, let us concentrate rather on explaining to human beings what we want
a computer to do.}''}{Donald E. Knuth}

This chapter describes the experimentation we conducted for
\correctingText{evaluating} the $\pi$SOD-M methodology. The chapter presents
three \correctingText{examples}. The first two \correctingText{examples} concern simplified
service-based applications that focus on the application's logic and explicit business rules. They show
how $\pi$SOD-M can be used for developing applications and exemplify the
consideration of non-functional aspects along the whole
development process. The last \correctingText{example}
(\textit{GesIMED}~\cite{valeriaThesis} application) \correctingText{aims to perform an
evaluation of our approach, analyzing the main differences and particularities
between SOD-M and $\pi$SOD-M.}

\newText{The objective of the experimentation was to evaluate the features and
concepts proposed by $\pi$SOD-M, considering the implementation of some
examples, and perform a comparison with SOD-M. This comparison focuses on
verifying which of those proposed aspects favored the development of
service-oriented applications in the presence of non-functional requirements. 
In this direction, our intention is: (i) analyze if it is possible to model
non-functional requirements from initial levels of abstraction in the
service-oriented development, (ii) and verify what are the advantages of using
the proposed approach, compared with the original proposal of SOD-M.}

% O objetivo principal deste capítulo é avaliar as características e conceitos
% adicionados em piSOD-M e fazer uma comparação com SOD-M, com a intenção de
% destacar que é possível modelar requisitos não funcionais desde níveis iniciais
% de abstração no desenvolvimento orientado a serviços.

%  used for
% validating the specification and generation of the models proposed in
% $\pi$SOD-M. The validation split the thesis into parts, each one can be
% validated individually. This parts are: (i) the modelling of each $\pi$SOD-M
% model, (ii) the model transformations and (iii) the results.
% 
%  
% The validation is performed through three case studies. The first case study
% aims to perform a qualitative evaluation of our approach, analyzing the main
% differences and particularities between SOD and $\pi$SOD-M. The others case
% studies aim to validate and verify the use of $\pi$SOD-M and its
% environment, analyzing the positive results and the benefits to 
% model non-functional requirements from early stage of development.

%The thesis' validation of the  follows from the sub-validations.

%\bigskip

% The \textit{IEEE Standard Computer Dictionary} \cite{ieeeDict} defines
% validation as \textit{the process of evaluating a system or component during or
% at the end of the development process to determine whether it satisfies specified
% requirements}. If we apply this definition to a methodology: \textit{Validation
% is the process of evaluating a method during or at the end of the development process to determine whether it
% satisfies specified requirements and eases the development of a system}. This
% chapter validates the methodology proposed in this thesis in order to verify
% whether it promotes a better development of reliable
% services composition applications.

% The steps during the thesis were analyzed by application of laboratory case
% studies with an iterative and incremental approach. Thus, while the $\pi$SOD-M
% concepts have been defined, the concepts application have been made in each case
% studies to verify its validity.

% The validation process has been performed in a continuous and
% progressive approach along the definition of $\pi$SOD-M.

This chapter is organized as follows: section
\ref{sec:toPublishMusic_inpractice} shows the \textit{To Listen Music}
\correctingText{example}, section \ref{sec:crimes_inpractice} introduces the
\correctingText{example} \textit{Crime Map} and section \ref{sec:siw_gesimed}
introduces the \textit{GesIMED} \correctingText{example} that we used for
validating the methodology. Section \ref{sec:leassonsLearned} presents the lessons learned
with the validation process. Finally, Section \ref{sec:conclusion_validation} concludes
the chapter and discusses on validation results.

% From strategy for
% validation, the process was carried out with the development resolution step as
% part of an iterative process and feedback cyclic, allowing solve problems of
% concepts and its application while the methodology was being defined. Thus, in
% each step of methodology the validation was applied iteratively
% on two different case studies.
% 
% The process defined in the research method validation was therefore not
% a linear process, but iterative and incremental, thus providing the opportunity
% to go strengthen the results and learning from each research step.  
% 
% Following, each developed case study is described, indicating both the needs
% identified, and new contributions emerged from them, which are also shown
% graphically, using the methodology models. The case studies resolution were
% performed using the proposed approach.

% \section{$\pi$SOD-M in Practice}
% \label{sec:execution}

% This section presents, in detail, the design and implementation of two case
% studies conducted to validate the method proposed in this
% thesis. Both case studies were developed using the models proposed by
% $\pi$SOD-M.

% Following all models will be developed for this case and then
% describe the main aspects of the implementation of each of them.

\section{Example 1: To Publish Music}
\label{sec:toPublishMusic_inpractice}

An organization wants to provide the service-based
 application that monitors the music a person is listening during some periods
 of time and sends the song title to this person's Twitter or Facebook accounts.
 Thus, this social network user will have her status synchronized in Twitter or
 Facebook. The scenario begins by get through the music service Spotify for
 fetching the user's musical status. The social network services are then
 contacted to modify the user's status with the music information. The user can
 also download a specific music, he can proceed with the download process, which
 includes the payment process.

The user can download a music, after having proceed with the
 payment via PayPal or credit card. All services are available via Spotify and
 this application needs to interact with Spotify users, so they can listen
 music, publish them on Facebook and Twitter, or buy music online. The scenario
 to monitor and publish songs has the User and Spotify actors.

The functionality of the application are: \textit{Search music;
Choose music; Download music; Listen Music; Buy music} and \textit{Publish music}.

 \subsection{$\pi$-UseCase Model}
\label{sec:usecase_topublishMusic}

Figure \ref{fig:publishMusic_usecase} shows the main part of the use case model
for the \textit{To Publish Music} application\footnote{The complete
\textit{$\pi$-UseCase} model can be found in appendix \ref{append:diagrams}.}.
The $\pi$-UseCase model describes the system functions and constraints. Figures
\ref{fig:toolpublishMusic_usecase} and \ref{fig:toolpublishMusic_usecase2}
detail the model and configuration properties for the described example
(actual view taken from the $\pi$SOD-M environment).
Three restrictions are specified on the use cases described by the
model, two on the \textit{buy music} use case and a restriction on the
\textit{pay} use case. These restrictions are value and business constraints.
\textit{Buy music} restrictions are related to secure connection and to
ensure that the user needs to have an account in the music network. They are
marked using \textit{\#\ldots\#}, figures \ref{fig:publishMusic_usecase},
\ref{fig:toolpublishMusic_usecase} and \ref{fig:toolpublishMusic_usecase2}.
Notice that those NFRs and constraints will be the base to define assertions and
contracts by the transformations of models.

\begin{figure}[ht!]    
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/piusecase/application/packages2.pdf}
\caption{\textit{To Publish Music} $\pi$-UseCase.}
\label{fig:publishMusic_usecase}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create_model-UseCase.pdf}
\caption{\textit{To Publish Music} $\pi$-UseCase Environment Detail.}
\label{fig:toolpublishMusic_usecase}
\end{figure}

Figures \ref{fig:toolpublishMusic_usecase} and
\ref{fig:toolpublishMusic_usecase2} present the update music model and show how
the properties are defined in the $\pi$SOD-M environment. This example describes
the properties over the \textit{update music} (item 4, figure
\ref{fig:toolpublishMusic_usecase}) use case and the restriction over the
authentication process (item 3, figure \ref{fig:toolpublishMusic_usecase}). To
update a music in a social network, the user must be authenticated.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-UseCase.pdf}
\caption{\textit{To Publish Music}$\pi$-UseCase Environment Detail.}
\label{fig:toolpublishMusic_usecase2}
\end{figure}

The modeling of the application properties using the environment $\pi$SOD-M
are the same presented in the $\pi$-UseCase model, as described in Figure
\ref{fig:toolpublishMusic_usecase2}. For example, the authentication restriction
 details the type of constraint (\textit{VALUE}), a \textit{description},
 \textit{name} and a \textit{non-functional attribute}
 (\textit{authentication}). These properties are described in all constraints
 (item 2, figure \ref{fig:toolpublishMusic_usecase2}).

Figure \ref{fig:servicesToPublishMusic} shows the service components that must
interact with the \textit{application} and the interaction among them. There are
four services that are invoked to produce results for the \textit{Publish Music} application:
\textit{Spotify, Facebook, Twitter} and \textit{Bank}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.8\textwidth]{chapters/validation/figs/toPublishMusic}
\caption{To Publish Music Services.}
\label{fig:servicesToPublishMusic}
\end{figure}

 \subsection{$\pi$-ServiceProcess Model}
\label{sec:serviceProcess_topublishMusic}

Figure \ref{fig:publishmusic_piserviceProcessToolModelProperties} shows the
\textit{$\pi$-ServiceProcess} diagram for the execution of activities based on
contracts. We present here the part of the diagram produced from the concepts of the figure
\ref{fig:toolpublishMusic_usecase2}. The complete \textit{$\pi$-ServiceProcess}
diagram of the application is shown in figure \ref{fig:serviceprocessContract}.
Notice that it is necessary to represent each {\sc Constraint} modelled in a {\sc Contract} based service activity diagram. Each constraint
described in the previous model is transformed into a contract in the
$\pi$-ServiceProcess model.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-ServiceProcess.pdf}
\caption{\textit{To Publish Music} $\pi$-ServiceProcess.}
\label{fig:publishmusic_piserviceProcessToolModelProperties}
\end{figure}

\textit{Publish Twitter} and \textit{publish Facebook} (figure
\ref{fig:publishmusic_piserviceProcessToolModelProperties}) have pre- and
post-conditions assertions that are grouped into a contract for each service:
(i) verify if the user data are correct; (ii) if the user is already logged in
Spotify; As post-condition, it ensures the complete transaction and
verifies whether the payment authorization notification was sent to the user and to Spotify.
There may be different restrictions depending on the services participating in
the publish process. The user must be logged in Spotify in order to access
Facebook, and Facebook must send a ``successful login'' acknowledge which is
verified by the post-condition. For publishing a message, Twitter imposes
pre-conditions on the length of the message and user data login. In contrast,
Facebook only sends a confirmation acknowledge. Items 4 and 5 in figure
\ref{fig:publishmusic_piserviceProcessToolModelProperties} show the
pre-conditions for ``\textit{publish Twitter}'' concerning \textit{(i)} the format
of the music file and (\textit{ii}) access authorization credentials, namely
login and password.

\correctingText{The development of the $\pi$-ServiceProcess refine
the constraints into contracts for the services as shown
in Figure \ref{fig:publishmusic_piserviceProcessToolModelProperties}.}

% , about the payment authorization. To \textit{update} services, depending of each
% service, there may be different restrictions.

%  The Facebook service requires that the user is already logged on
%  Spotify and these data are the same as Facebook. As post-condition, it ensures
%  that the Facebook service send a notification of success. The Twitter service
%   requires pre-conditions over the message length and user privacy data, while
%   to update the Facebook is a necessary precondition and a confirmation notice
%   as post-condition. As a precondition for ``\textit{publish twitter}'' (itens
%   4 and 5, figure \ref{fig:publishmusic_piserviceProcessToolModelProperties}) it is necessary that
%   \textit{(i)} the music format is correct and (\textit{ii})  the twitter login and password is correct for the update.

 \subsection{$\pi$-ServiceComposition Model}
\label{sec:serviceComposition_topublishMusic}

 There are four external {\sc Business
Collaborators} in this application: \textit{Bank, Spotify, Twitter} and
\textit{Facebook} (items 3, figure
\ref{fig:topublishMusic_piserviceCompositionToolModel}). Figure \ref{fig:topublishMusic_piserviceCompositionToolModel} shows the application
services that consist of a set of functions: \textit{search music, select
music, buy music, download music, listen music} and \textit{publish music}. The
\textit{publish music} action calls two service collaborators (Facebook and
Twitter), and the \textit{buy music} action calls two functions of the Bank
service collaborator.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create-model-Serviceconposition.pdf}
\caption{\textit{To Publish Music} $\pi$-ServiceComposition Model.}
\label{fig:topublishMusic_piserviceCompositionToolModel}
\end{figure}

The \textit{To Publish Music} $\pi$-ServiceComposition model defines three
policies: \textit{transactionPolicy} (item 4, figure
\ref{fig:topublishMusic_piserviceCompositionToolModel}), \textit{httpAuthPolicy}
(item 2) and \textit{authPolicy} (item 6). The \textit{trasactionPolicy}
verifies the Bank data restrictions over the payment and confirmation process.
The \textit{httpAuthPolicy} policy specifies rules over the
Facebook function for the authentication and notification processes. These rules
are executed during the publish massage function request. The
\textit{authPolicy} policy implements the open authentication protocol
required for accessing the Twitter service.

From the \textit{To Publish Music} rules and policies it is possible to model
and associate non-functional properties to services' compositions,
\textit{e.g.} the integration of information retrieved from different
social network services, automatic generation of an integrated view of the
operations executed in different social networks or for providing security in
the communication channel when the payment service is called.

From this stage, the \textit{$\pi$-PEWS} model (PSM) depicted in figures
\ref{fig:pewsRepresentation} and \ref{fig:contractRepresentation} can be
generated.

%  \subsubsection{$\pi$-PEWS Model}
% \label{sec:pews_topublishMusic}

% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.90\textwidth]{chapters/implementation/figs/create_model-PEWSProperties.pdf}
% \caption{$\pi$-PEWS Model Properties in $\pi$SOD-M Eclipse Plugin.} 
% \label{fig:piPEWSToolModelProperties}
% \end{figure}

% \section{Results}
% \label{sec:results}

\section{Example 2: Crime Map}
\label{sec:crimes_inpractice}

The \textit{Crime Map} \correctingText{example} addresses the design and
implementation of a system that deals with statistical data on criminal activities. The system works
in conjunction with third-party services, in order to process a search supplied
by a user. The system must provide a public utility service to inform
citizens about the real crime situation in a particular location or region.
Making use of third party services, the application must provide a statistics crime portal. The result depends on user queries and upon a user
demand results can be presented on a map. The system also interacts with
micro-blogging, for looking up messages containing specific crime
information. The presentation of the data concerning criminal activities is
provided by the \textit{Police service}. The presentation of a locality map
crime is made by the \textit{Google Maps service} and messages presentation on
the micro-blogging is done by the \textit{Twitter service}. The system also has
a \textit{Post service}, for searching an address location that can be then
presented in a map by Google maps. The scenarios to query data and crime
statistics have the following actors: \textit{(i) User, (ii) Map service, (iii) Post service, (iv) Micro-blogging
service} and \textit{(v) Police service}, to query the criminal activities
(figure \ref{fig:usecaseCrimeMap_total}).

Considering the application functions, the crimes may be searched by types,
\textit{e.g.} \textit{theft, murder, kidnapping}, etc. The search can also be
made by city' district or region, for example, \textit{center, west, south,
north} or \textit{east}, and it can process search temporal criteria, for
example, crimes of the previous day of the last weeks or months. Finally, the
service can also process conjunctive queries, \textit{e.g.}, \textit{the number of kidnapping of a specific district in the last month}.

Whenever a new crime is published by the police, the system should post a
 message on Twitter stating the new crime, with a specific hashtag,
 \textit{e.g.} \textit{\#crime043\_kidnapping\_center\_05-08-2012}. The hashtag
 structure contains the crime number, type, local and the day it happens. From
 this, users can comment about the crime and these comment will appear in the
 application interface. All users must have an account to use the application.

 The application functionalities are: \textit{Choose search type of crime; see
 crime information; Receive information on the chosen query; See map; List crimes;}
 and \textit{Comment about a crime.}

 Non-functional requirements for this application include: \textit{queries
 response performance, crime information validation}, \textit{crime
 information presented in a map} and \textit{ensuring crimes data security.}

\subsection{$\pi$-UseCase Model}
\label{sec:usecase_crimesMap}

% The modeling of the system starts with the $\pi$-UseCase model. This involves
% determining the services identified in the list of requirements, which final
% consumers identified in the list of business services, and which of the will be
% end consumers or service users. Thus, it is possible to identify the modeling
% elements that are part of the use case model.

 Figure \ref{fig:usecaseCrimeMap_total} presents the Crime Map $\pi$-UseCase
model. This model includes examples of \textit{business, value}
and \textit{exceptional behaviours} constraints. The application accesses
data from third party services to provide the user an easy way to visualize new
information, these services are: \textit{Map, Police, Postal} and
\textit{Microb-logging Services}.

The user interacts with the application that will perform the services
for obtain and publish details about the crimes. The services are modeled as
actors (they are external applications). Each service has a set of functions
that can be performed. The $\pi$-UseCase model describes both the external services
and the functionality of the application (see figure
\ref{fig:usecaseCrimeMap_total}).

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase}
\caption{Crime Map $\pi$-UseCase.}
\label{fig:usecaseCrimeMap_total}
\end{figure}

The user can search for crimes and view its occurrences in a map. The search may be performed by \textit{type of
crime, region} or \textit{period}. The user can also perform a search that
combines these criteria. If the application's user does not choose any
option, the search will retrieve up to 30 recent occurrences. If the location
maps service is unavailable or there is no response from it after 5 seconds,
only the crimes information will be presented, without a map view.

Figure \ref{fig:usecaseCrimeMap_total} also
presents the $\pi$-UseCase model tree representation generated by the $\pi$SOD-M
environment model .Figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2} detail the
use cases and constraints presented in Figure
\ref{fig:usecaseCrimeMap_total}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase1}
\caption{\textit{Search crime} $\pi$-UseCase Detail.}
\label{fig:usecaseCrimeMap1}
\end{figure}

The \textit{search crimes} use case (Figure \ref{fig:usecaseCrimeMap1})
requests the \textit{Police} service to verify the crimes according to the
parameters supplied by the user. The restriction of this use case depends on
the parameters selected by the user. If the user chooses one or several types of
specific crime's type, the system will present those crimes that have more than
five occurrences. For \textit{search by address}, the address must comply with
the restrictions of the police service, the research can be done by street,
neighborhood, city or district. \textit{Search by period} will provide only
occurrences within the last six months.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase2}
\caption{\textit{See crime information} $\pi$-UseCase Detail.}
\label{fig:usecaseCrimeMap2}
\end{figure}

Figure \ref{fig:usecaseCrimeMap2} describes the design of the use cases:
\textit{see crime information, see statistics, show map} and \textit{share
information}. To share information of a crime, it is necessary to access the
Twitter micro-blogging service to proceed with the authentication and posting.
The service requires the user id and password, however the
application must ensure the User privacy information while performing
authentication. To see statistics of crimes, the application should process a
larger volume of data to generate statistical results. Thus, it is required to
provide resources to ensure the data processing performance. The restriction on
map presentation is over response time and accessibility of the \textit{Postal}
and \textit{Google maps} services. The response time limit is 5 seconds. If the
map is not processed, only the information of the crime are presented. Figures
\ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2} also show the
properties used in the $\pi$SOD-M environment and the equivalent concepts in the
model.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.8\textwidth]{chapters/validation/figs/crimesMap}
\caption{Crime Map Services.}
\label{fig:servicesCrimeMap}
\end{figure}

After the identification the various functions that are required by the system
to perform the business services, the $\pi$-ServiceProcess model is used to
represent the workflow necessary to perform a service within the system.

\subsection{$\pi$-ServiceProcess Model}
\label{sec:serviceprocess_crimesMap}

The $\pi$-ServiceProcess model is used to represent the workflow necessary to
perform the system service. The service processes execution is described in two
steps. Figures \ref{fig:serviceProcessCrimeMap} and \ref{fig:serviceProcessCrimeMap2} show the
flow of functions execution described in the $\pi$-UseCase models shown
in Figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}. Each
diagram can represent different business services (see figures
\ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}).

The diagrams of figures \ref{fig:serviceProcessCrimeMap} and
\ref{fig:serviceProcessCrimeMap2} were obtained by applying the
\textit{$\pi$-UseCase2$\pi$-ServiceProcess} transformation to the diagrams of
figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}, respectively.

%  Thus, it makes easier to represent the performance of services. It is also possible to describe the
% process in a large workflow, but for applications running multiple
% services, it is better to have smaller processes to better represent the
% service execution. Thus we present the services business through the different
% processes as presented in Figures \ref{fig:usecaseCrimeMap1} and
% \ref{fig:usecaseCrimeMap2}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceProcess}
\caption{\textit{Crime information} $\pi$-ServiceProcess.}
\label{fig:serviceProcessCrimeMap}
\end{figure}

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceProcess_crime1}
\caption{\textit{Crime information}
$\pi$-ServiceProcess Environment.}
\label{fig:toolserviceProcessCrimeMap}
\end{figure}

Each constraint of figures \ref{fig:serviceProcessCrimeMap} and
\ref{fig:serviceProcessCrimeMap2} are transformed into assertions. For example,
the assertions of address format,  crime types and time are restrictions over 
the action \textit{search crime}, being one pre-condition and two
post-conditions. These restrictions form the contract on the \textit{search
crime} function. Similarly, for viewing maps, if the answer to the request
takes longer than 5 seconds (5000 milliseconds), the map display is suspended and only 
the information is presented. Each assertion in this $\pi$-ServiceProcess model
stems from the $\pi$-UseCase model constraints. These restrictions are shown
in Figure \ref{fig:serviceProcessCrimeMap}. Figure
\ref{fig:toolserviceProcessCrimeMap} presents the equivalent model
generated by the methodology transformation in the $\pi$SOD-M environment for
the process detailed in Figure \ref{fig:serviceProcessCrimeMap}. Each node
defines the process element and its properties.

In Figure \ref{fig:toolserviceProcessCrimeMap}, \textit{i\_node} corresponds
to the initial process node. {\sc Actions} are grouped into  {\sc Service
Activities}, such as \textit{choose crimes property} action that is part of
the \textit{search crime} service activity. A {\sc Control Flow} is an edge that
links two nodes, for example \textit{cf1} connect the initial node
\textit{i\_node} (source) with the \textit{choose crimes property} action
(target). Figure \ref{fig:toolserviceProcessCrimeMap} also presents the
assertions described in Figure \ref{fig:serviceProcessCrimeMap} grouped into 
{\sc Contracts}. The contracts are: \textit{searchCrimeContract} and
\textit{seeCrimeInformationContract}. These contracts are related with the
actions \textit{search crime} and \textit{see crimes information}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceProcess2}
\caption{\textit{See crime statistic and share information} $\pi$-ServiceProcess
Detail.}
\label{fig:serviceProcessCrimeMap2}
\end{figure}

Figure \ref{fig:serviceProcessCrimeMap2} presents the assertions over the
 \textit{see statistics} and \textit{share information} actions. Both have 2 
pre-conditions. To \textit{share information} on Twitter it is necessary
to verify the format of crime information (140 characters) and the Twitter id
and password, for authentication. Regarding the \textit{see statistics} action, there is a
business restriction over data volume, and a value restriction over the crime
information format. Thus, the presentation of the crimes statistics must be done
after the contract verification. Figure
\ref{fig:toolserviceProcessCrimeMap2} presents the equivalent model
generated by the methodology transformation in the $\pi$SOD-M environment for
the process detailed in Figure \ref{fig:serviceProcessCrimeMap2}. Each node
defines the process element and its properties. Figure
\ref{fig:toolserviceProcessCrimeMap2} also presents the assertions grouped
into three contracts that are related with its specific actions. The contracts
are: \textit{seeStatisticsContract}, \textit{seeCrimeInformationContract} and
\textit{shareInformationContract}. The contracts are related with the action
\textit{see statistics}, \textit{see crimes information} and \textit{share
information}, respectively.

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceProcess_crime2}
\caption{\textit{Crime statistic and share information}
$\pi$-ServiceComposition Environment.}
\label{fig:toolserviceProcessCrimeMap2}
\end{figure}

These models provide an overview of the business processes from the requirements
described in the $\pi$-UseCase model. These models also provide a more detailed
view of the execution of business processes and the possible
interaction with external services. The description of this interaction is the
result of the refinement of these models ($\pi$-ServiceProcess models) through
the $\pi$-ServiceComposition model.

\subsection{$\pi$-ServiceComposition Model}
\label{sec:servicecomposition_crimesMap}

Figures \ref{fig:serviceCompositionCrimeMap} and
\ref{fig:serviceCompositionCrimeMap2} show the $\pi$-ServiceComposition models,
which represent the service composition processes of each
business service and additional indication of which are members of the business
that execute the action. Actions are derived from service activities,
identifying the set of actions that are necessary for the completion of each service.

The main partition expresses the \textit{Application} execution
(\textit{External false}) that represents the general process. Actions have the equivalent
service relation, which realize the external service, such as \textit{Police}, \textit{Google Maps} or
\textit{Twitter} services. They are expressed by a {\sc Business Collaborator}.
Both, Figure \ref{fig:serviceCompositionCrimeMap} and
\ref{fig:serviceCompositionCrimeMap2} present two external {\sc Business
Collaborators} describing the application services that are invoked. These
$\pi$-ServiceComposition models refine the $\pi$-ServiceProcess models, matching
action with real service functions that may be executed by the system
application.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceComposition}
\caption{\textit{Crime information}
$\pi$-ServiceComposition.}
\label{fig:serviceCompositionCrimeMap} 
\end{figure}

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceCOmposition_crime1}
\caption{\textit{Crime information}
$\pi$-ServiceComposition Environment.}
\label{fig:toolserviceCompositionCrimeMap}
\end{figure}

For each $\pi$-ServiceComposition model, the contracts described in the
$\pi$-ServiceProcess model are grouped in policies. Figure
\ref{fig:serviceCompositionCrimeMap} describes the \textit{performancePolicy}
and \textit{reliabilityPolicy} policies, both associated with a service activity
and its actions. It is important to highlight that all policies are applied to the
entire actions associated with a service activity, for example, the
\textit{performancePolicy} verifies the request time for presenting the Google
maps. If the time exceeds 5000 milliseconds, the maps call is ignored and only
the crime information is presented. The \textit{reliabilityPolicy} is
associated with crime data verification. There are three rules over crime
address format, types of crime and the period of time in which it happens. The
crime presentation may obey the pre- and post-condition described in this policy. Figure
\ref{fig:toolserviceCompositionCrimeMap} presents the $\pi$SOD-M tool
description for the model described in Figure
\ref{fig:serviceCompositionCrimeMap}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceComposition2}
\caption{\textit{Crime statistic and share information}
$\pi$-ServiceComposition.}
\label{fig:serviceCompositionCrimeMap2}
\end{figure}

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceCOmposition_crime2}
\caption{\textit{See crime statistic and share information}
$\pi$-ServiceComposition Environment Detail.}
\label{fig:toolserviceCompositionCrimeMap2}
\end{figure}

Figure \ref{fig:serviceCompositionCrimeMap2} refines the $\pi$-ServiceProcess
model described in Figure \ref{fig:serviceProcessCrimeMap2} and Figure
\ref{fig:toolserviceCompositionCrimeMap2} presents the $\pi$SOD-M tool
description for the model described in Figure
\ref{fig:serviceCompositionCrimeMap2}. This model details the policies for
\textit{crime statistic} and \textit{share information} services. The
policies are \textit{statisticPolicy} and \textit{authenticationPolicy},
respectively.

These models provide an overview of the external services and its composition
from the actions described in the $\pi$-ServiceProcess model. These models also
provide a more detailed view of the system restrictions, applying policies
over external services. This model refines the $\pi$-ServiceProcess models
detailing the Business Collaborators that are expressed as external services,
and grouping contract into policies.

% From these composition models, next
% step is generate the specification system code in $\pi$-PEWS.

% \subsubsection{$\pi$-PEWS Model}
% \label{sec:pews_crimesMap}

\begin{figure}[ht!]
\tiny
\centering

\begin{lstlisting}[label=list:crimeInformationPEWS,caption=pi-PEWS
Specification: Crime information. ]

1  ns police =  "http:\\www.pm.rn.gov.br/service"
2  ns googleMaps =  "http://maps.googleapis.com/maps/api/"

4  alias chooseCrimesProperties = portType/chooseCrimesProperties in police
5  alias searchCrimes = portType/searchCrimes in police
6  alias seeMap = portType/showMap in googleMaps
7  alias seeCrimesInformation = portType/getCrimeDetail in police

9 (chooseCrimesProperties . searchCrimes)* . seeMap . seeCrimesInformation

11 def contract searchCrimesContract{
12  isAppliedTo: searchCrimes;    
13  requires: addressFormat == ?? 
14     (onFailureDo: call(chooseCrimesProperties) );  
15    ensures : !(getCrimes(types).length > 5) ==> false &&
16              result[i].period < 1800
17       (onFailureDo: show(getCrimesInformation())); 
18 }

20 def contract showMapContract{
21  isAppliedTo: seeMap;    
22    ensures : serviceResponse.getTime() < 5000
23       (onFailureDo: skip); 
24 }
\end{lstlisting}
\label{fig:crimeInformationPEWS}  
\end{figure}

The result of the transformation of this model to the \textit{$\pi$-PEWS} model
is gives in the listing \ref{list:crimeInformationPEWS} and
\ref{list:crimeStatisticsPEWS}.

\begin{figure}[ht!]
\tiny
\centering
\begin{lstlisting}[label=list:crimeStatisticsPEWS,caption=pi-PEWS Specification: Crime
statistic and share information. ]

1  ns police =  "http:\\www.pm.rn.gov.br/service"
2  ns twitter =  "https://dev.twitter.com/docs/api/"

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
\urldef{\mailsa}\path|{\ldots,|
\urldef{\mailsb}\path|\ldots,|
\urldef{\mailsc}\path|\ldots}@\ldots.com|    
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Web Services Composition in the presence of Non-Functional Properties: A
MDD Approach}

% a short form should be given in case it is too long for the running head
%\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Pl\'acido A. Souza Neto \and Martin A. Musicante \and\\%
%
 Genoveva Vargas-Solar\and Umberto Costa\and Val\'eria de Castro\\}
%
%\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{\ldots\\
\mailsa\\
\mailsb\\
\mailsc}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

%\toctitle{Lecture Notes in Computer Science}
%\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
\ldots.
\keywords{\ldots}
\end{abstract}

\section{Introduction}
\ldots

\section{A methodology for web services com\-po\-si\-tions}

In this section we present $\pi$SOD-M. $\pi$SOD-M is a MDA (Model Driven
Architecture) based methodology. It provides a environment for building service compositions considering
their non-functional requirements. $\pi$SOD-M extends the SOD-M \cite{valeriaThesis} method by adding
the concept of \textit{Policy} \cite{Espinosa-OviedoVZC09,Espinosa-Oviedo2011a}
for representing NFR associated to service-based applications. $\pi$SOD-M also
proposes the generation of a set of models at different abstraction levels, as
well as transformations between these models.

$\pi$SOD-M's models represent both the cross-cutting aspects of the application
being modelled, as well as the constraints associated to services. The systems
cross-cutting concerns affect functional concerns, such as availability;
recovery; and persistence aspects. Constraints are restrictions that must be
respected during the execution of the application, for example the fact that a
service requires an authentication for executing system functions.

$\pi$-SOD-M supports the construction of service-oriented applications that implement business processes.
Therefore, it proposes a development process based on the definition of models
(instances of the meta-modes) and transformations for semi-automatically
generating different levels of abstraction that describe a service-oriented
application from abstract platform independent views (CIM and PIM level) to
platform dependent views and the PSM and implementation levels. We extended the
Business and Service views of the SOD-M method \cite{CastroMV11}. The Business
view defines  the concepts for modeling a business process, while the Service
view defines the concepts  for designing services compositions. Our methodology
introduces concepts (e.g. NF-requirement, constraint, assertion, contract,
policy) in the Policy view for describing constraints associated to services and
non-functional properties associated to service processes.

\section{Modeling web services com\-po\-si\-tions with
$\pi$SOD-M}
